# More than two levels {#sec-regression-more}

![](https://img.shields.io/badge/Area-R-green)

@sec-regression-cat showed you how to fit a regression model with a categorical predictor. We modelled reaction times as a function of word type (real or nonce). The categorical predictor `IsWord` (word type) was coded with so-called treatment contrasts: the model's intercept is the mean of the first level and the "slope" is the difference between the second level and the first.

```{r}
#| label: eu-vot

library(tidyverse)

eu_vot <- read_csv("data/egurtzegi2020/eu_vot.csv")
eu_vot
```

```{r}
#| label: eu-vot-three

eu_vot <- eu_vot |> 
  mutate(
    phonation = case_when(
      phone %in% c("p", "t", "k") ~ "voiceless",
      phone %in% c("b", "d", "g") ~ "voiced",
      phone %in% c("ph", "th", "kh") ~ "aspirated"
    ),
    VOT = VOT * 1000
  )

```

```{r}
#| label: fig-eu-vot-dens

eu_vot |> 
  drop_na(phonation) |> 
  ggplot(aes(VOT, fill = phonation)) +
  geom_density(alpha = 0.5)

```

```{r}
#| label: fig-eu-vot-jit

eu_vot |> 
  drop_na(phonation) |> 
  ggplot(aes(phonation, VOT)) +
  geom_jitter(alpha = 0.5, width = 0.2)

```

```{r}
#| label: vot_bm

library(brms)

vot_bm <- brm(
  VOT ~ 0 + phonation,
  family = gaussian,
  data = eu_vot,
  seed = 6725,
  file = "cache/ch-regression-more-vot_bm"
)

```

`Warning: Rows containing NAs were excluded from the model.`

```{r}
#| label: vot-bm-summ
summary(vot_bm)
```
